## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <= 5) {
  print(a)
} else {
  print("It's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 10

if (a <= 5) {
  print('small')
} else if (a > 5 & a <= 10) {
  print("medium")
} else if (a > 10 & a <= 15) {
  print("large")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 6

if (a <= 5) {
  if (a < 4 ){}
    print("less than 4!")
    
} else {
  print('hi')
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){
  # paste is a concatenation function
  print(paste(value,"is greater than", value -1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1)

#change the first,third, and sixth elements to the value 6 

for (i in c(1,3,6)){
  a_vector[i] <-  6

}
a_vector
```

We can loop over more than just lists:

```{r}
mat <-  matrix(1, nrow = 2,ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row,col] <-  0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <-  0 
while (b < 10){
  b <-  b+2
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
our_function()
```

An example:

```{r}
# start w a vector of temperatures 
test_temp <- c(16.54056,26.98654,32.456745,8.6547,28.45674)

# function to convert from C to F: 
c_to_f <-  function(dat){
  temp_f <- dat*(9/5) +32
  return (temp_f)
}

c_to_f(26)
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_f <-  c_to_f(test_temp)

data_f
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
yuh <-matrix(c(1,1,1,2,2,2), nrow <- 2)
# 1 is for rows, 2 is for columns 
apply(yuh,1,sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_tempm <- matrix(c(16.54056,26.98654,32.456745,8.6547,28.45674),nrow = 2)

apply(test_tempm,1,c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #1: Create a function that takes an integer as an argument and returns "Even" for even numbers or "Odd" for odd numbers.
    even_or_odd <- function(n) {
      if (n%%2 == 0){
        return('Even')
      } else{
        return('Odd')
      }
    }
    even_or_odd(2)
    ```

    ```{r}
    #2:We need a function that can transform a number (integer) into a string.
    number_to_string <- function(n) {
      n  <- as.character(n)
      return (n)
    }
    ```

    ```{r}
    #3: Finish the solution so that it returns the sum of all the multiples of 3 or 5 below the number passed in.
    solution <- function(number){
      bucket <- 0
      for(i in 1:number-1){
        if (i %% 3 == 0 | i %% 5 == 0){
          bucket <- bucket + i
        }
      }
    return(bucket)
    } 

    ```

    ```{r}
    #4: You get an array of numbers, return the sum of all of the positives ones.
    positive_sum <- function(vec){
      pos_sum <- 0
      for (i in vec){
        if (i>0){
          pos_sum <- pos_sum + i 
        }
      } 
    return(pos_sum)
    }
    ```

    ```{r}
    #5: In this kata, you are asked to square every digit of a number and concatenate them.For example, if we run 9119 through the function, 811181 will come out, because 92 is 81 and 12 is 1. (81-1-1-81)
    square_digits <- function(num){
      sqstr <-  ""
     x <- as.numeric(strsplit(as.character(num), "")[[1]])
     for (i in x){
       square <-  i*i
       sqstr <- paste(sqstr,square,sep="")
     }
      return(as.numeric(sqstr))
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

```{r}
#check one number only
  fizzbuzz <- function(i){
        if(i %% 3 == 0 & i %% 5 == 0){
          return('fizzbuzz')
        }else if(i %% 3 == 0 ){
          return('fizz')
        }else if(i %% 5 == 0 ){
          return('buzz')
        }else{
          return(i)
        }
        }
  
fizzbuzz(3)
fizzbuzz(5)
fizzbuzz(7)
fizzbuzz(30)
```

```{r}

#with a list of numbers
  n <- seq(1,20)
  fizzbuzzl <- function(num){
        for(i in num){
            if(i %% 3 == 0 & i %% 5 == 0){
              print('fizzbuzz')
            }else if(i %% 3 == 0 ){
              print('fizz')
            }else if(i %% 5 == 0 ){
              print('buzz')
            }else{
            print(i)
          }
          }
        }
        fizzbuzzl(n)
```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    stair <- function(n){
      for (i in 1:n){
        print(paste(replicate(i,"X"),collapse =""))
      }
    }
    stair(5)
    ```
